---
title: "assignment 3"
author: "Beimnet Taye"
date: "2024-03-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)

```

```{r se}
library(tidyverse)
library(R2jags)
library(knitr)
library(kableExtra)
library(coda)
library(foreign)
load("CVD_data.Rdata")

# Extract data elements from data frame
cvd <- CVD.data$cvd # Outcome
cursmoke <- CVD.data$cursmoke # Exposure (smoking status)
age.c <- as.numeric(scale(CVD.data$age))
male <- as.integer(CVD.data$sex=="male")

X.educ <- model.matrix(~-1 + factor(educ), data=CVD.data)
educ1 <- X.educ[,1] # Unused in our analysis (reference category)
educ2 <- X.educ[,2]
educ3 <- X.educ[,3]
educ4 <- X.educ[,4]
```

```{r jags}
model.posterior <- function() {
for (i in 1:N) {

  
logit(pi[i]) <- b[1] + b[2] * cursmoke[i] + b[3] * age.c[i] + b[4] * male[i] + b[5] * educ2[i] + b[6] * educ3[i] + b[7] * educ4[i];
  
cvd[i] ~ dbin(pi[i], 1); # Distribution of outcome
}
# PRIORS ON BETAS
for (j in 1:Nx){
b[j] ~ dnorm(mu[j], tau[j]); # Independent normal priors
OR[j] <- exp(b[j]); # Calculate the odds ratios
}
}
```

```{r jagscons}

# Constants to be passed in
N <- length(cvd) # Number of observations to loop over
Nx <- 7 # Number of parameters (w/ intercept)

n.iter <- 10000 # Number of iterations to run (total)
# Parameters on the priors:

mu <- rep(0,Nx)

tau <-  rep(1/1000, Nx)
# List of data elements to pass in
data.list <- list("N", "Nx", # Model constants (# obs, # vars)
"cvd", "age.c", # Variable names (next 3 lines)
"male", "cursmoke",
"educ2", "educ3", "educ4",
"mu","tau") # Hyperparameters
# List of parameters to keep track of:
parameters.model <- c("b", "OR")
# Function to randomly generate initial values for each chain:
inits.model <- function() {list (b=rnorm(Nx, 0, sd=.5))}

```

```{r run}
set.seed(123)
jags.samples <- jags(data=data.list,
model.file=model.posterior,
inits=inits.model,
parameters.to.save=parameters.model,
n.iter=n.iter, n.chains=3,
quiet = T)
# print(jags.samples,digits=4)
```

```{r diag}
mcmc.samples <- as.mcmc(jags.samples) # Converts samples to "MCMC object"
# # for diagnostics
# # Traceplot and density plots for regression coefficients
# # code will save to PDF in current directory.
# # Execute "plot" commands only to plot to screen.
# pdf("Traceplot_LogisticReg.pdf") # Write what comes next to PDF file
# plot(mcmc.samples[1][,1:4]) # For Chain 1, beta1-4
# plot(mcmc.samples[1][,5:8]) # For Chain 1, beta5-7 and deviance
# dev.off() # Stop writing to the PDF file
# 
# 
# # Autocorrelation plots for the regression coefficients
# pdf("ACF_LogisticReg.pdf")
# par(omi=c(.25,.25,.25,.25)) # Create an outer margin (room for title)
# autocorr.plot(mcmc.samples[1][,1:7]) # For chain 1
# title("Chain 1", outer=T) # Place title in outer margin of page
# autocorr.plot(mcmc.samples[2][,1:7]) # For chain 2 (optional)
# title("Chain 2", outer=T)
# autocorr.plot(mcmc.samples[3][,1:7]) # For chain 3 (optional)
# title("Chain 3", outer=T)
# dev.off()

# geweke.diag(mcmc.samples[,1:7])
```

```{r inform1}
mu[2] <- log(2)
set.seed(123)
jags.samples.inform1 <- jags(data=data.list,
model.file=model.posterior,
inits=inits.model,
parameters.to.save=parameters.model,
n.iter=n.iter, n.chains=3,
quiet = T)
# print(jags.samples.inform1,digits=4)
```

```{r inform2}

sd.prior <- (log(2.67) - log(1.5))/(2*1.96) 
tau[2] <- (1/sd.prior)^2
set.seed(123)
jags.samples.inform2 <- jags(data=data.list,
model.file=model.posterior,
inits=inits.model,
parameters.to.save=parameters.model,
n.iter=n.iter, n.chains=3,
quiet = T)
# print(jags.samples.inform2,digits=4)

```
\newpage
# Q1

```{r q1}

noninform <- jags.samples[[2]]

inform1 <- jags.samples.inform1[[2]]

inform2 <- jags.samples.inform2[[2]]

noninform <- noninform$summary[c(2:7),c(3,5,7)] %>% 
  round(2) %>% 
  as.data.frame() %>% 
  mutate(est = paste0(`50%`," (",`2.5%`,", ", `97.5%`,")")) %>% 
  select(est)

rownames(noninform) <- c("Current smoker (vs. non)", "Age (per year increase)", "Male sex (vs. female)", "High school education (vs. <
HS)", "Some college (vs. < HS)", "College+ (vs. < HS)")

inform1 <- inform1$summary[c(2:7),c(3,5,7)] %>% 
  round(2) %>% 
  as.data.frame() %>% 
  mutate(est = paste0(`50%`," (",`2.5%`,", ", `97.5%`,")")) %>% 
  select(est)

rownames(inform1) <- c("Current smoker (vs. non)", "Age (per year increase)", "Male sex (vs. female)", "High school education (vs. <
HS)", "Some college (vs. < HS)", "College+ (vs. < HS)")

inform2 <- inform2$summary[c(2:7),c(3,5,7)] %>% 
  round(2) %>% 
  as.data.frame() %>% 
  mutate(est = paste0(`50%`," (",`2.5%`,", ", `97.5%`,")")) %>% 
  select(est)

rownames(inform2) <- c("Current smoker (vs. non)", "Age (per year increase)", "Male sex (vs. female)", "High school education (vs. <
HS)", "Some college (vs. < HS)", "College+ (vs. < HS)")

estimates <- cbind(noninform, inform1,inform2)

colnames(estimates) <- c("Vague Priors", "Informative Prior 1", "Informative Prior 2")

estimates %>% 
  kable(booktabs =T, col.names = c("Vague Prior", paste0("Informative Prior 1", footnote_marker_symbol(1)), paste0("Informative Prior 2", footnote_marker_symbol(2))),
  escape = F) %>% 
  footnote(symbol = c("Prior mean for OR of current smoking=2, prior variance of log-OR=1000.", "Prior mean for OR of current smoking=2, prior variance of log-OR= 0.02.")) %>% 
  kable_styling(latex_options = c("HOLD_position", "striped"))
  
```

\newpage
# Q2

```{r q2}

p1 <- c(log(2) - 1.96*sqrt(1000), log(2) + 1.96*sqrt(1000))

p2 <- c(log(2) - 1.96*sqrt(0.02), log(2) + 1.96*sqrt(0.02))

```

The prior 95% interval for Informative Prior 1 is: $log(2) \pm 1.96*\sqrt{1000}$ which is equal to: $(-61.3,62.7)$. This is a wide prior for the beta coefficient for smoking, especially when compared to the corresponding interval for Informative Prior 2, making it less informative than that of Informative Prior 2.

\newpage
# Q3

Informative Prior 2 had more of an effect of dragging the smoking OR closer to 2 since the precision is larger for informative prior 2 resulting in prior 2 to influence the posterior distribution more.

\newpage
# Q4
Based on the traceplots it seems to model converged nicely for all parameters measured. Every parameter has a hairy caterpillar like trace around a horizontal line indicating good convergence. The density plots also seem pretty uni-modal indicating good convergence. The autocorrelation plots for all parameters indicate that sample auto correlation drops off pretty quickly indicating good model mixing and convergence. 

\newpage
# Q5

```{r}

geweke.diag(mcmc.samples[,1:7])
```


From the results of the Geweke test it seems the 5th beta in the first chain and the second beta in the third chain failed to converge as their Z test statistics were greater than 1.96 in magnitude, indicating that the mean of the first 10% of the chain differed significantly from the mean of the final 50% of the chain. This means that the sampler varied/wandered too much when sampling. 
